A weakly structured stem for human origins in Africa

### 2023-03-04 paper.tex
Revisions to address Editor's comments

Evidence for archaic hominin admixture in Eurasia has bolstered the plausibility of 
archaic hominin admixture having also occurred in Africa. Previous work that sought to 
explain patterns of polymorphism inconsistent with a single-origin model thus focused on 
archaic hominin admixture as an alternative model, by referring to additional (or ``ghost'') 
branches required to fit the data as ``archaic'' 
\citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz,
Durvasula2020-td} and assuming (or inferring) deep divergences. These perspectives have 
oriented interpretations of both genomic (e.g., selection \cite{Wall2019-ao}) and fossil 
(e.g., evolution of \emph{Homo sapiens} \cite{Grun2020-su}) data. Here, we have shown that 
a model of a weakly structured stem better captures the apparently inconsistent patterns 
of polymorphisms.

@@

Third, multiple studies have shown a correspondence between phenotypic
differentiation, usually assessed with measurements of the cranium, and genetic
differentiation among human populations and between humans and Neanderthals
\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb} (see also SI
Section~\ref{sec:morpho_diff}). This correspondence potentially allows
predictions of our model to be related to the fossil record. The fossil record
of Africa is sparse during the time period of the stems, but of the fossils
that date to this time period, some are fairly similar overall in morphology to
contemporary humans (e.g., Omo 1 from Omo Kibish, Ethiopia
\citep{Day1969-rh,Vidal2022-qe}), others are similar in some morphological
features to contemporary humans (e.g., Irhoud 1 from Jebel Irhoud, Morocco
\citep{Hublin2017-cq,Richter2017-zu}), and others are different enough in
morphology to have been assigned to species other than \emph{Homo sapiens}
(e.g., DH1 from Dinaledi, South Africa \citep{Berger2015-bq,Dirks2017-uk}). If,
as our model predicts, the genetic differences between the stems were
comparable to those among contemporary human populations, the most
morphologically divergent fossils are unlikely to represent branches that
contributed appreciably to contemporary human ancestries.

>>>>

Evidence for archaic hominin admixture in Eurasia has bolstered the plausibility of 
archaic hominin admixture having also occurred in Africa. Previous work that sought to 
explain patterns of polymorphism inconsistent with a single-origin model thus focused on 
archaic hominin admixture as an alternative model, by referring to additional (or ``ghost'') 
branches required to fit the data as ``archaic'' 
\citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz,
Durvasula2020-td} and assuming (or inferring) deep divergences. These perspectives have 
oriented interpretations of both genomic (e.g., selection \cite{Wall2019-ao}) and fossil 
(e.g., evolution of early \emph{Homo sapiens} \cite{Grun2020-su}) data. Here, we have shown 
that a model of a weakly structured stem better captures the apparently inconsistent patterns 
of polymorphisms.

@@

Third, multiple studies have shown a correspondence between phenotypic differentiation, 
usually assessed with measurements of the cranium, and genetic differentiation among 
human populations and between humans and Neanderthals \citep{Relethford1994-mh,
Weaver2008-ho,Von_Cramon-Taubadel2009-zb} (see also SI Section~\ref{sec:morpho_diff}). 
This correspondence potentially allows predictions of our model to be related to the 
fossil record. Some \emph{Homo sapiens} fossils, from Iho Eleru, Nigeria (13ka) 
\citep{Bergstrom2021-iw} \textcolor{red}{ADD REFERENCE Harvati et al. 2011 ``The Later 
Stone Age Calvaria from Iwo Eleru, Nigeria: Morphology and Chronology''}, Ishango, 
Democratic Republic of Congo (25-20ka) \textcolor{red}{ADD REFERENCE Crevecoeur et al. 
2016 ``Late Stone Age Human Remains from Ishango (Democratic Republic of Congo): New 
Insights on Late Pleistocene Modern Human Diversity in Africa''}, and Nazlet Khater, 
Egypt (40-35ka) \textcolor{red}{ADD REFERENCE Crevecoeur 2012 ``The Upper 
Paleolithic Human Remains of Nazlet Khater 2 (Egypt) and Past Modern Human Diversity''}, 
have morphological features that may reflect gene flow from archaic hominins. If 
this interpretation is correct, these fossils are unlikely to represent populations 
ancestral to contemporary human populations, because we would have expected to detect 
this gene flow and we do not. In our multiple-merger model, we do infer late gene flow 
from one of the stems into West African populations at 25ka, but this is not admixture 
with an archaic hominin lineage. The fossil record of Africa is sparse during the time 
period of the stems, but of the fossils that date to this time period, some are fairly 
similar overall in morphology to contemporary humans (e.g., Omo 1 from Omo Kibish, Ethiopia 
\citep{Day1969-rh,Vidal2022-qe}), others are similar in some morphological features to 
contemporary humans (e.g., Irhoud 1 from Jebel Irhoud, Morocco 
\citep{Hublin2017-cq,Richter2017-zu}), and others are different enough in morphology to 
have been assigned to species other than \emph{Homo sapiens} (e.g., DH1 from Dinaledi, 
South Africa \citep{Berger2015-bq,Dirks2017-uk}). If, as our model predicts, the genetic 
differences between the stems were comparable to those among contemporary human populations, 
the most morphologically divergent fossils are unlikely to represent branches that 
contributed appreciably to contemporary human ancestries.
